---
title: "Chapter 12, Logical "
format: html
---

In this chapter, you’ll learn tools for working with logical vectors. **Logical vectors** are the simplest type of vector because each element can only be one of three possible values: TRUE, FALSE, and NA. It’s relatively rare to find logical vectors in your raw data, but you’ll create and manipulate them in the course of almost every analysis.

We’ll begin by discussing the most common way of creating logical vectors: with *numeric comparisons*. Then you’ll learn about how you can use Boolean algebra to combine different logical vectors, as well as some useful summaries. We’ll finish off with if_else() and case_when(), two useful functions for making conditional changes powered by logical vectors.

------------------------------------------------------------------------

## Prerequisites

```{r libraries}
library(tidyverse)
library(nycflights13)
```

Let's also make up some dummy data with c():
```{r dummy data}
x <- c(1, 2, 3, 5, 7, 11, 13)
x * 2
```

Just remember that any manipulation we do to a free-floating vector, you can do to a variable inside a data frame with mutate() and friends.
```{r dummy data tibble}
df <- tibble(x)
df |> 
  mutate(y = x * 2)
```

------------------------------------------------------------------------

## Comparisons

